---
title: "Assignment 2"
subtitle: 'Life Insurance Mathematics'
author: "Asparuh Rangelov (13576534), Denis Kolarik (14047063)"
date: "2023-09-15"
output: pdf_document
---
```{r}
library(RColorBrewer)
cols = brewer.pal(9,"Set1")
col<-cols[1:30]
```

We import mortality data for Germany:

```{r}
Data = read.table('/Users/asparuhrangelov/Downloads/Germany_A2.txt', header = TRUE)

```
Q2 A

With the code below we plot the force of mortality for an (x) years olds, in relation between time and age, for different time periods:

```{r}
for (i in 1:9){
  Data[,i] = as.numeric(Data[,i])
}
Data$mux = -1*log(1-Data$qx)
```
Above, we created a new variable 'mux', that approximates the force of mortality. We used the approximation

$$
\mu_x \approx -\ln(1-d_x)
$$

```{r}

with(subset(Data, Year == 1990),
     plot(Age, log(mux),type = "l", ylim = c(-10,0), main = "Force of mortality for Germay: 1990-2020",
          xlab = "Age x", ylab = expression(paste("ln(Force of mortality)")),col=col[1], lwd = 2))
with(subset(Data, Year == 1990 + 10),
     lines(Age, log(mux) ,col=col[2], lwd = 2))
with(subset(Data, Year == 1990 + 20),
     lines(Age, log(mux) ,col=col[3], lwd = 2))
with(subset(Data, Year == 1990 + 30),
     lines(Age, log(mux) ,col=col[4], lwd = 2))
legend(3,0,legend=c("t=1990","t=2000","t=2010",'t=2020'),
       col=col[1:4], lty=1,cex=0.9, lwd = 1.5)
```

Question 2 B

For this subquestion, we plot the logarithm of the force of mortality, as previously calculated, for different years. We vary this across ages.
```{r}
with(subset(Data, Age == 10),
     plot(Year, log(mux) ,type = "l", ylim = c(-10,0.5), main = 
            "Force of mortality versus year, for different ages",
          xlab = "Year", ylab = expression(paste("ln(Force of mortality)")),col=col[1], lwd = 2))
with(subset(Data, Age == 20),
     lines(Year, log(mux) ,col=col[2], lwd = 2))
with(subset(Data, Age == 30),
     lines(Year, log(mux) ,col=col[3], lwd = 2))
with(subset(Data, Age == 40),
     lines(Year, log(mux) ,col=col[4], lwd = 2))
with(subset(Data, Age == 50),
     lines(Year, log(mux) ,col=col[5], lwd = 2))
with(subset(Data, Age == 60),
     lines(Year, log(mux) ,col=col[6], lwd = 2))
with(subset(Data, Age == 70),
     lines(Year, log(mux) ,col=col[7], lwd = 2))
with(subset(Data, Age == 80),
     lines(Year, log(mux) ,col=col[8], lwd = 2))
legend(1989.14,0,legend=c("Age = 10","Age = 20","Age = 30",'Age = 40', 'Age = 50',
                    'Age = 60', 'Age = 70', 'Age = 80'), horiz = TRUE,
         col=col[1:8], lty=1,cex=0.58, lwd = 1.5)


```

For the specified period (1990-2020), there seems to be not big variation for the force of mortality across all ages. Nonetheless, there is a general trend for the force of mortality to decrease with time.

Question 2 C

For this question, we define the cummulative distribution of lx by using the 'cumprod' command. Furthermore, we condition on the years 1990-2020 (increasing by a factor of 5):

```{r}
Data_1990<-subset(Data,Year==1990)
qx_1990<-as.numeric(Data_1990$qx)
px_1990<-1-qx_1990

p0_1990<- cumprod(px_1990[1:length(px_1990)])
Age_1990<-Data_1990$Age
plot(Age_1990,p0_1990,type = "l",main = expression(paste("Survival Function\ ","S" [0],","[t],"(x)")) ,
     ylim = c(0,1),xlab = "Age x",ylab = 'Survival Probability of a newborn' ,col = col[1], lwd = 1.5)


years<-seq(1995,2020, by = 5)

for (i in 1:length(years)) {
  Data_i<-subset(Data,Year==years[i])
  qx_i<-as.numeric(Data_i$qx)
  px_i<-1-qx_i
  p0_i<- cumprod(px_i[1:length(px_i)])
  Age_i<-Data_i$Age
  lines(Age_i,p0_i,col = col[i+1], lwd = 1.5)
  }

legend(10,0.8,legend=c("t=1990","t=1995","t=2000","t=2005","t=2010", 't=2015', 't=2020'),
       col=col[1:7], lty=1,cex=1, lwd = 1.5)

```

The trend is clear: survival probabilities tend to improve with time, with the most recent survival function being on the left.

Question 2 D

For this question, we plot the number of people, out of 100,000, that die immediately after surviving to age x. 

For this, we use the product (lx * muX), which is the probability that someone survives x years and dies immediately after that.

For muX we use the approximation mx $\approx -1\cdot \ln(1-d_{x})$.

```{r}

Data_1990<-subset(Data,Year==1990)
qx_1990 = as.numeric(Data_1990$qx)
mx_1990 = as.numeric(Data_1990$mx)
lx_1990 = as.numeric(Data_1990$lx)

y = lx_1990 * -1*log(1-qx_1990)

Age_1990<-Data_1990$Age
plot(Age_1990,y,type = "l", main = "Number of people (of 100K) that die after reaching an age x", 
     ylim = c(0,5000),xlab = "Age x",ylab = 'lx * muX' ,col = col[1], lwd = 1.5)

years = seq(1995,2020,by = 5)

for (i in 1:length(years)) {
  Age_i<-Data_i$Age
  Data_i<-subset(Data,Year==years[i])
  qx_i = as.numeric(Data_i$qx)
  mx_i = as.numeric(Data_i$mx)
  lx_i = as.numeric(Data_i$lx)
  
  y_i = lx_i * -1*log(1-qx_i)
  
  lines(Age_i, y_i, col = col[i+1], lwd = 1.5)
  }

legend(10,4500,legend=c("t=1990","t=1995","t=2000","t=2005","t=2010", 't=2015', 't=2020'),
       col=col[1:7], lty=1,cex=1, lwd = 1.5)

```

The graph confirms our intuition, namely that most people arrive only after reaching 60, with the peak being around 80.

The same trend for improvement in life expectancy we mentioned previously is also visible here: the more recent probabilities of surviving are pointwise better than the older ones.



